---
title: "Curated NEON datasets"
output: github_document
---

Contains scripts for downloading and cleaning data, and the resulting data files. 
Metadata for original and curated datasets are in this README. 

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

### 1. Plant cover

#### Original data

- **Plant presence and percent cover** dataset 
- Product ID *DP1.10058.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.10058.001)
- Summary: Plant cover for each species of plant was estimated in six 1m2 subplots within 400m2 plots, where plant cover was percent of subplot ground covered as viewed from above. Each site has around 30 plots, with sites distributed across the USA. Plant cover was taken multiple times per year over multiple years, depending on the site. 
- Additional useful information
  - Some plants have vouchers/tissues collected that may be useful for genetic analyses
  - The only data for plant height is `heightPlantOver300cm`, which indicates whether plants are taller than 9.8 feet

#### File structure

- `plant_cover` folder
  - Scripts
    - `curate_data.R` cleans up data
    - `summarize_data.R` creates summary stats, tables, and figures of data
  - Derived data and figures
    - `plant_cover.csv` is curated data
    - All `.png` images result from `summarize_data.R` and are displayed below

#### Curated data details

Columns: 

- `species`: species identification
- `lat`: latitude of plot (decimal degrees)
- `lon`: longitude of plot (decimal degrees)
- `sitename`: site, plot, and subplot info combined in format `sitecode_plotID_subplotID`; e.g., `DSNY_DSNY_017_32.4.1` is site DSNY, plot 017, subplot 32.4.1
- `date`: date of end of sampling in format YYYY-MM-DD
- `canopy_cover`: amount of ground covered by that species in 1m2 area (%)
- `uid`: unique identifier for each record as assigned by NEON

Summary figures and stats: 

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
plant_cover <- read.csv("plant_cover/plant_cover.csv")
```

**Locations**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sites_plots <- plant_cover %>% 
  separate(sitename, sep = "_", into = c("site", "also_site", "plot", "subplot")) %>% 
  group_by(site) %>% 
  summarise(count = n_distinct(plot)) %>% 
  rename(Site = site, Plots = count)
```

- `r nrow(sites_plots)` sites with `r sum(sites_plots$Plots)` total plots
- Coordinates correspond to plot, not subplot
- Map of plot locations: 

```{r, echo=FALSE}
map_background <- map_data("state") 
         
ggplot() +
  geom_polygon(data = map_background, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") + 
  geom_point(data = plant_cover, aes(x = lon, y = lat), 
             color = "blue", shape = 4) +
  labs(x = "", y = "") +
  theme_classic()
```

- Figure of number of plots per site, ordered by number of plots:  

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(sites_plots) +
  geom_col(aes(x = reorder(Site, -Plots), y = Plots)) +
  xlab("Site") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
```

**Taxonomy**

- `r nrow(plant_cover)` records for `r length(unique(plant_cover$species))` species
- Table of the 20 species with the most records and their number of occurrences: 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
species_counts <- plant_cover %>%
  group_by(species) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:20) %>%
  rename(Species = species, Occurrences = count)
kable(species_counts, format = "markdown")
```

**Time**

- Records taken on `r length(unique(as.Date(plant_cover$date)))` days from `r min(as.Date(plant_cover$date))` to `r max(as.Date(plant_cover$date))`
- Plot of number of records per day across entire time range: 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dates <- plant_cover %>% 
  select(date) %>% 
  mutate(date = as.Date(date)) %>% 
  group_by(date) %>% 
  summarize(count = n()) %>% 
  rename(Date = date, Records = count)

ggplot() +
  geom_col(data = dates, aes(x = Date, y = Records), color = "black", fill = "black") +
  theme_classic()
```

### 2. Phenology measurements

#### Original data

- **Plant phenology observations** dataset 
- Product ID *DP1.10055.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.10055.001)
- Summary: At each site for multiple years, phenophase measurements for ~100 individual plants made up to multiple times a week depending on phenology activity. 

### 3. Phenology images

#### Original data

- **Phenology images** dataset 
- Product ID *DP1.00033.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.00033.001)
  - Data stored on [PhenoCam](https://phenocam.sr.unh.edu/webcam/about/) website [here](https://phenocam.sr.unh.edu/webcam/network/search/?sitename=&type=&primary_vegtype=&dominant_species=&active=unknown&fluxdata=unknown&group=neon); probably have to be downloaded individually by site? 
- Summary: Images (RGB and IR) taken from tops of towers at each site every 15 minutes, available for most sites back to early 2017. 

### 4. Mammal diversity

#### Original data

- **Small mammal box trapping** dataset 
- Product ID *DP1.10072.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.10072.001)
- Summary: For all sites starting in 2016, individual small mammals were trapped and identified 4-6 times per year. 

### 5. Bird diversity

#### Original data

- **Breeding landbird point counts** dataset 
- Product ID *DP1.10003.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.10003.001)
- Summary: Species identifications of individual non-large and non-aquatic birds from a point. Each site has 5-10 grids with 9 points per grid (45-90 points), with counts occurring once or twice a year since 2017 for most sites. 

### 6. Mosquito diversity

#### Original data

- **Mosquitoes sampled from CO2 traps** dataset 
- Product ID *DP1.10043.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.10043.001)
- Summary: All sites were sampled, starting in 2016 for some, once to twice a week at 10 plots per site (note: sampling design changed somewhat in 2018). Subsamples from traps were identified to taxon and includes abundance numbers. 

### 7. Beetle diversity

#### Original data

- **Ground beetles sampled from pitfall traps** dataset 
- Product ID *DP1.10022.001*
- [Data portal link](https://data.neonscience.org/data-products/DP1.10022.001)
- Summary: All sites were sampled, starting in 2016 for most, twice a week at 10 plots per site. Individuals from traps were identified to species. 
